---
title: "Bayesian Data Analysis"
subtitle: "What Does My Model Assume?"
# For author options see : 
# https://quarto.org/docs/authoring/front-matter.html#authors-and-affiliations
# NB: multiple authors can be added here.
author:
  - name:
      given: Joshua
      family: Wilson Black
    email: joshua.black@canterbury.ac.nz
    orcid: 0000-0002-8272-5763
    affiliation: 
      - "Te Kāhui Roro Reo | New Zealand Institute of Language, Brain and Behaviour"
      - "Te Whare Wānanga o Waitaha | University of Canterbury"
format:
  revealjs:
    theme: [custom.scss]
    incremental: true
    logo: images/NZILBB-small.svg
    template-partials:
      - title-slide.html
    title-slide-attributes:
      # If you have Marsden funding, change image name to `nzilbb-uc-marsden.svg`
      data-background-image: images/nzilbb-uc.svg
      # First number controls the horizontal position, second controls vertical.
      data-background-position: '50% 5%'
      # Controls size of image relative to width of the slide.
      data-background-size: 70%
    embed-resources: false
    include-in-header:
      - text: |
          <link rel="icon" type="image/png" sizes="32x32" href=".//images/fav.png" />
bibliography: 
  - grateful-refs.bib
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
knitr:
  opts_chunk: 
    dev: "png"
    dev.args: 
      bg: "transparent"
---

```{r}
#| echo: false
# This is a place to load packages in the background (i.e., this block won't 
# apear on the slides because `echo` is false.)
library(tidyverse)
library(here)

library(brms)

theme_set(
  theme_bw(base_size=20) +
    theme(
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA),
      legend.background = element_rect(fill = "transparent", colour = NA)
    )
)

qb2 <- read_rds(here('data', 'session_1.rds'))

trap_sub <- qb2 |> 
  filter(
    school %in% c(
      "Avonside Girls' High School", 
      "St Margaret's College"
    )
  ) |> 
  select(
    age, school, F1_lob2, part_id
  )
```

# Overview

## Overview

1. What is a prior?
2. Prior predictive tests
3. Sensitivity analysis

# What's a prior?

## The prior {.smaller}

- Bayesian models start with a 'prior' distribution.
- The prior says what results are plausible [before]{.red} we see the data.
- A prior might be (roughly):
    1. **uninformative**: doesn't push towards any result,
    2. **weakly informative**: pushes towards plausible results but can 
    be overcome by sufficient data,
    3. **strongly informative**: pushes towards plausible results, essentially
    ruling out others.
- Strong v. weak is a continuum.
- These are the three options relevant for reporting.

## What actually is Bayes' Theorem?

::: {.fragment .big}

$p(\theta|e) = \frac{p(e|\theta)p(\theta)}{p(e)}$

:::

- $p(e|\theta)$ - The [likelihood]{.red} of the evidence given the hypothesis.
- $p(\theta)$ - The [prior]{.red} probability of the hypothesis.
- $p(\theta|e)$ = The [posterior]{.red} probability of the hypothesis, given the
evidence.
- The prior and likelihood should be interpreted together.

::: footer
In practice, you can ignore $p(e)$.

:::

## The prior is just part of the model 

- The prior captures information we have before the data (see Gelman etc.)
- Treat it as just another part of the model.
- Practical tests of priors:
    - Convince yourself that it's compatible with any plausible outcome 
    (via a [prior predictive test]{.red}).
    - Check how much your choice influences your result 
    (via a [prior predictive test]{.red}).
- A terminological niggle: you can change your prior, **after** you see your data!

## Probability distributions...

::: r-stack

:::: {.fragment .fade-in-then-out .r-fit-text}

Cauchy

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

Dirichlet

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

Poisson

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

Gaussian

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

Beta

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

Gamma

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

Wishart

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

Student's $t$

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

There's no way around them...

::::

:::: {.fragment .fade-in-then-out .r-fit-text}

Borrow, test and modify

::::

:::

## Uninformative 

- Uninformative priors let the data totally dominate.
- In practice are very rare (although sometimes used in `brms` defaults)

::: fragment

```r
default_prior(
    F1_lob2 ~ school,  
    data = trap_sub
)
```
:::

## Uninformative

```{r}
#| echo: false
default_prior(
    F1_lob2 ~ school,  
    data = trap_sub
  ) |> 
  tibble() |> 
  select(prior, class, coef) |> 
  mutate(
    prior = if_else(prior == "", "(flat)", prior)
  ) |> 
  knitr::kable()
```

- `(flat)` indicates an uninformative prior.
- `student_t(...)`, on intercept and variance, is weakly informative. 


## Weakly informative

::: fragment
```{r}
#| output-location: fragment
rstudent_t(n=1000, 3, 0.4, 2.5) |> 
  hist()
``` 

:::

## Weakly informative (cont.) {.smaller}

::: columns

:::: {.column width="25%"}
- The intercept prior is centred on the mean F1,
- The variance is also estimated from the data.
- It's compatible with a very wide range of plausible results.
::::

:::: {.column width="75%"}
```{r}
#| echo: false
rstudent_t(n=1000, 3, 0.4, 2.5) |> 
  hist()
``` 
::::

:::

- But *not* with, e.g., an intercept of 2000.
- Weakly informative priors are typically the best choice.
- **NB:** `brms` defaults use the data to estimate the prior.

## Weakly informative (cont. 2)

- Informativeness is a continuum.
- Let's be more informative...

::: fragment
```{r}
trap_prior <- c(
  prior(student_t(3, 0, 2), class = "Intercept"),
  prior(normal(0, 2), class = "sigma"),
  prior(normal(0, 0.5), class = "b")
)
```

:::

::: fragment
```{r}
#| cache: true
#| output: false
trap_fit_b <- brm(
  F1_lob2 ~ school,  
  data = trap_sub,
  prior = trap_prior
)
```

:::

## Weakly informative (cont. 3) {.smaller}

::: fragment

> We fit a Bayesian linear regression with weakly informative priors using the `brms` pakage. The intercept, representing the mean F1 for Avonside students, has a weakly informative Student t distribution (df=3, mean=0, scale=2) with moderately
heavy atails. The difference between Avonside and 
St Margarets has a normal (mean=0, sd=0.5) prior, which expresses the expectation that the students come from the same dialect and that their mean 
[trap]{.smallcaps} F1 is very unlikely to be more than 1 point different in Lobanov normalised space.
The prior on residual standard deviation is normal (mean=0, sd=2), accomodating a wide range of variation with Lobanov normalised formant values.
:::

## Strongly informative

- Our weakly informative prior was based on readily available information about, e.g.,
Lobanov normalised formant values.
- We might want to add new data to our existing model.
- We might have some other source on information.
- This can be done using 'strongly informative' priors.

## Strongly informative (cont. 2)

- Here's a strong **and false** prior.

::: fragment
```{r}
#| cache: true
#| output: false
bad_prior <- c(
  prior(student_t(3, 0, 2), class = "Intercept"),
  prior(normal(0, 2), class = "sigma"),
  prior(normal(0, 0.01), class = "b")
)
trap_fit_s <- brm(
  F1_lob2 ~ school,  
  data = trap_sub,
  prior = bad_prior
)
```
:::


## Strongly informative (cont. 3) {.smaller}

::: fragment

```{r}
#| output-location: fragment
rnorm(1000, mean=0, sd=0.01) |> 
  hist()
```
:::

- We're not allowing differences larger than about
0.03 in normalised space.
- Presumably this is not actually detectable.

## Strongly informative (cont. 4) {.smaller}

::: fragment
```{r}
fixef(trap_fit_s)
```
:::

::: fragment

vs.

:::

::: fragment
```{r}
fixef(trap_fit_b)
```
:::

- Our prior expectation on the size of the difference
between schools has messed up both the intercept (Avondside) _and_ the St Margaret's coefficients.

## A rule

::: {.fragment .r-fit-text}
Explicitly specify your prior!
:::

- Don't use the default without thinking
- If you do decide to use the default, put it in your code explicitly.
    - i.e., not 'we used the default prior',
    - rather, 'use used an uninformative prior for all coefficients, with a weakly informative [etc.]' 

# Prior predictive checks

## The idea

- Last week we drew from the posterior distribution.
- We can also draw from the prior and see what comes
out (using the likelihood).
- Easy to do with `brms`.
- If the values are implausibly wide, we can narrow the prior (and vice versa)

## Implementation

::: fragment
```{r}
#| cache: true
#| output: false
#| code-line-numbers: |5
trap_ppc <- brm(
  F1_lob2 ~ school,  
  data = trap_sub,
  prior = trap_prior,
  sample_prior = "only"
)
```
:::

## Sampling

::: fragment
```{r}
library(tidybayes)
prior_draws <- trap_ppc |>
  predicted_draws(newdata = trap_sub, ndraws = 100)
```

:::

## Plotting

::: fragment

```r
prior_draws |> 
  ggplot(
    aes(
      x = school,
      y = .prediction
    )
  ) +
  geom_boxplot()
```
:::

## Plotting (cont. 2) {.smaller}

::: columns
:::: {.column width="70%"}

```{r}
#| echo: false
prior_draws |> 
  ggplot(
    aes(
      x = school,
      y = .prediction
    )
  ) +
  geom_violin()
```

::::

:::: {.column width="30%"}
- Bulk sitting in plausible values
- Outliers are a bit excessive 
- We could tighten this up if we wanted
- In more complex models, this is more useful!
::::

:::

- Note that this wouldn't reveal problem in the strong prior above.
- Just looking at a N(0, 0.01) distribution would be enough.

# Sensitivity analysis

## The idea

- Try some other priors, see how much it affects the results.
- I'd always do a prior predictive check, but only do a sensitivity analysis if I were worried.

## Automate {.smaller}

- The `priorsense` package is an option [@priorsense2023].

::: fragment
```{r}
#| output-location: fragment
library(priorsense)
powerscale_sensitivity(trap_fit_b)
```
:::



## Automate (cont. 2) {.smaller}

::: fragment

```{r}
#| output-location: fragment
powerscale_plot_dens(trap_fit_b)
```
:::

## Automate (cont. 3) {.smaller}

::: fragment

```{r}
#| output-location: fragment
powerscale_sensitivity(trap_fit_s)
```
:::

# Summary {.smaller}

## Summary

1. What is a prior?
    - incl. uninformative, weakly informative, or strongly informative.
2. Prior predictive tests
    - i.e. what possibilities is my prior consistent with?
3. Sensitivity analysis
    - i.e. what are the consequences of changing the prior for my results?

## Now and next time

- Download GitHub repository: <https://github.com/nzilbb/ws-bayes-3>
- Work through `analysis.R`
- Next time: Is this thing healthy?

# References

```{r}
#| echo: false
grateful::nocite_references(
  grateful::cite_packages(output = "citekeys", out.dir = here())
)
```


::: refs

:::
